MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-25 response_2_reviewers.txt

     We have complied with and/or answered all the reviewer comments, and have also gone through the ms again carefully and made other edits for a few corrections and many clarifications (changes in red).  The following detailed responses explain our reasoning, and quote the changes.

@@

Reviewer 1:
• Could you devise a figure or table that more directly or comprehensively compares the older (penultimate) rotamer library to the new (ultimate) you describe here

-->  In the supplement we added a table that compares the outlier count in the unfiltered dataset when evaluated using the Top500 vs. the Top8000 contours (see table 2 in supplement) and made a reference to that table in Section 4.2. We have also added a figure comparing the contours of the remaining 2-dimensional residue types not already in figure 2 (see Figure S1 in the supplement). Along with this we added references to Figure S1 in 4.2 and the caption of Figure 2 (See red text). Also at the beginning of Results we now compare the current number of named rotamers (212) with the number in our 2000 paper (153).

@@

• Should “chain-level” be expanded in the subheading to the section “Chain-level filters” to “Main-chain” or “backbone conformation”  or “overall quality filters” ?

-->  The subheading has been changed to "Overall Chain-Level Filters", and several small changes have been made in the preceding section to set readers up to realize that "chain" here means a PDB chain, not a polypeptide backbone, and that it is a minor shift from "file-level" filters and contrasts with residue-level filters.

• The following sentence on page 7 made no real sense “This entire set of datasets is therefore nicknamed the ”Top8000” of structural biology’s upper class.“

-->  We have now dropped this playful analogy to 19th Century British and American society's "Upper Ten Thousand", and the sentence now reads simply:

@@

* Section 5.1 should be a single well organized paragraph, not a collection of 5 single sentence paragraphs.  I’m also not completely sure this section (and the designation of ultimate) is completely necessary. It is quite clear you have made a contribution, it is perhaps less clear that we can be absolutely sure we will not want to iterate rotamer library development one or more times after this.

-->  It would be extremely difficult to make this section readable and its logic understandable as a single paragraph.  Instead, we have made it an explicit list of separate arguments and have rewritten it for better clarity and continuity, which is indeed quite helpful:

“Calling this set of rotamer-library distributions ”ultimate” is a claim that requires both explanation and justification. 16 years after our ”penultimate” rotamers, we had to confront the issue of whether or not MolProbity’s line of validation-focused rotamer distributions had reached its ultimate stage. We decided that indeed it had, from three separate lines of argument, reflected in the wording of this work’s title.

(1) We claim it as ultimate only for the multi-dimensional χ distributions used to validate sidechain models, such as done by the MolProbity website and in PHENIX. The rotamers presented here are not necessarily ultimate or even appropriate for other purposes.

@@

(2) The changes from penultimate to ultimate distributions are fairly minor in quantitative terms, even after 16 years. The importance of the new system lies in its ability to support a 3-level evaluation, to identify very rare rotamers, and to specify rotamer-dependent bond-angle deviations. A million reliably-modeled residues of data now provide fully adequate sampling for the basic task of robustly locating smooth contours that separate favored (98%), allowed, and outlier regions. The fundamental χ distributions would be unlikely to change much at all if they were redetermined in the future, and we plan to keep them freely available on GitHub for testing, for use within other software systems, and for spinoffs such as approximation by differentiable functions.

(3) What we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and computing power, and driven by the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together [8], including allowance for the influence of secondary structure and local motifs. Thus our ”ultimate” claim asserts the position that MolProbity’s type of rotamer evaluation should no longer be updated but should evolve into something better.
”

-->  The more important issue is the word "ultimate". Note that even the title says these are MOLPROBITY'S ultimate rotamer distributions (not everyone's), and for model validation, not for everything.  It is indeed quite likely that someone else will iterate rotamer development -- but we don't intend to, in anything like the same form.  We feel the paper makes our case adequately and prominently (in title, abstract, and discussion), and that "ultimate" can be treated as a meaningful and memorable, but not too dead-serious, designation alongside the previous “penultimate” library.

@@

"The p, t, m nomenclature was adopted in [1] and in MolProbity [Davis 2004], to give a single letter for use in rotamer strings, and because the more common g+, t, g- terminology was, in 2000, still being assigned inconsistently (see discussion in [1])”

• References to side chain building could include the PDB_REDO tool "Side_Aide" and the Lego_Side_Chain options from O.

-->  The following citation to SideAide was added in the introduction when talking about where rotamer libraries are used (brackets enclosing this citation are in red):

R. P. Joosten, K. Joosten, G. N Murshudov, A. Perrakis. PDB_REDO: constructive validation, more than just looking for errors. Acta Cryst D 68: 484-496, 2012.

### 2016-02-25 discussion.tex
Edits in response2 & several parts of ms

\textcolor{changecolor}{(1) 
We claim it as ultimate only for the multi-dimensional $\chi$ distributions used to validate sidechain models, such as done by the MolProbity website and in \textit{PHENIX}. The rotamers presented here are of course not necessarily ultimate or even appropriate for other purposes.}

@@

\textcolor{changecolor}{(3)  W}hat we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and computing power, and \textcolor{changecolor}{driven by} the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together \cite{JSR_theplot_2013}, including allowance for the influence of secondary structure and local motifs. Thus our "ultimate" claim asserts the position that \textcolor{changecolor}{MolProbity's type of }rotamer evaluation should no longer be updated but should evolve into something \textcolor{changecolor}{that treats covariation simultaneously}.

>>>>

\textcolor{changecolor}{(1) 
We claim it as ultimate only for the multi-dimensional $\chi$ distributions used to validate sidechain models, such as done by the MolProbity website and in \textit{PHENIX}. The rotamers presented here are not necessarily ultimate or even appropriate for other purposes.}

@@

\textcolor{changecolor}{(3)  W}hat we do expect will improve in the future is a redefinition of conformational validation, enabled by expansion of data and computing power, and \textcolor{changecolor}{driven by} the needs of structure determination at lower resolution. Rather than doing separate Ramachandran and rotamer evaluation, we should move toward analyzing all backbone and sidechain torsional dimensions together \cite{JSR_theplot_2013}, including allowance for the influence of secondary structure and local motifs. Thus our "ultimate" claim asserts the position that \textcolor{changecolor}{MolProbity's type of }rotamer evaluation should no longer be updated but should evolve into something \textcolor{changecolor}{better}.

### 2016-02-25 results.tex
Edits in response2 & several parts of ms

\section{Results}
In the Top8000 system there are a total of 212 named rotamers which reach above the 0.3\% outlier cutoff and \textcolor{changecolor}{passed our manual examination. Clusters with $\le$ 8 datapoints were manually examined -- two Lys clusters lacked sufficient density to justify their inclusion}. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see Section \ref{sec_md:chi_stats}), which we feel represents the clusters better than either the modal or the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S3 describes the frequency and count of each rotamer in each residue type and Tables S4-S21 give central dihedral and covalent angle values for each rotamer. These numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in Section \ref{sec_md:contours}.

>>>>

\section{Results}
In the Top8000 system there are a total of \textcolor{changecolor}{212 named rotamer clusters, compared with 153 in our 2000 paper [1]}. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see Section \ref{sec_md:chi_stats}), which we feel represents the clusters better than either the modal or the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S3 describes the frequency and count of each rotamer in each residue type and Tables S4-S21 give central dihedral and covalent angle values for each rotamer. These numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in Section \ref{sec_md:contours}.
